Guarded Recursion in Agda via Sized Type

### 2019-02-27 code-oneclock-paper/types-abstract/abstract.tex
started types abstract

@@

Guarded recursion \cite{Nakano}
is a different approach where the type system is enhanced with a
modality, called ``later'' and written \IC{▻}, encoding time delay in
types. The later modality comes with a general fixpoint combinator for
programming with productive recursive functions and allows the
specification of guarded recursive types.
These types can be used in combination with clock quantification to define  coinductive types \cite{atkey2013productive,bizjak2016guarded,BahrGM17}. Currently there is no
implementation of a calculus for guarded recursion.
\remove{, \eg the type of guarded
streams satisfying the isomorphism \F{Str} \Ar{A} \F{≃} \Ar{A} \F{×}
\IC{▻} (\F{Str} \Ar{A}). The tail of a guarded stream is available only
one time step ahead from now, as evidenced by the application of the
later modality to \F{Str} \Ar{A}. Coinductive types can be constructed
from guarded recursive types in type theories that additionally
support clock variables and universal quantification over them
\cite{atkey2013productive,BahrGM17}. Currently there is no
implementation of a calculus for guarded recursion.}
%%  nor guarded recursion
%% appears as a selectable feature in a proof assistant.

@@

Utilizing techniques for representing ``type theory in type theory''
\cite{AltenkirchK16,Chapman09}, we present an Agda formalization of
the syntax of a simple type theory for guarded recursion. This object
language, which we call \GTT, is a variant of Atkey and McBride's type
system for productive coprogramming \cite{atkey2013productive} in
which the clock context can contain at most one clock. The types of
\GTT\ include the aforementioned \IC{▻} modality and a \IC{□} modality,
a nameless analogue of Atkey and McBride's universal clock
quantification. Clouston \etal \cite{CloustonBGB15} studied a
guarded variant of lambda calculus extended with a \IC{□} operation,
which they call ``constant''.  Our object calculus differs from
theirs in that our judgments are indexed by a clock context, which can
be empty or containing exactly one clock. 
%In this respect \GTT\ is closer in spirit to Atkey and McBride's
%language.
The design of \GTT\ has the benefit of allowing a more appealing introduction rule for
\IC{□} than Clouston et al.'s.

@@

This paper only include the essential parts of our Agda
formalization. The full code is available at
\url{https://github.com/niccoloveltri/agda-gtt}. The formalization
uses Agda 2.5.4.2 and Agda standard library 0.16. The paper is
extracted from a literate Agda file, which implies that all the
displayed code has passed Agda's type and termination checker.

The material is organized in the following way.
In \Cref{sec:prelim}, we discuss the metatheory and we give an
overview of sized types.
In \Cref{sec:syntax}, we introduce the syntax of \GTT, our object type theory for guarded recursion.
Subsequently, we give a categorical semantics.
In \Cref{sec:presheaf_sem}, we show how to implement presheaves over
sizes and how to model the fragment of
\GTT\ corresponding to simply typed lambda calculus.
In \Cref{sec:guarded}, we discuss the semantics of the guarded
recursive and coinductive features.
In \Cref{sec:consistent}, we prove the object language sound \wrt the
categorical model,
%describe how to formally interpret the syntax into the model, 
which in turn entails the consistency of the syntax.
Finally, in \Cref{sec:concl}, we draw conclusions and suggest  future
directions of work.

%% give the Kripke semantics
%%\GTT.
%%For that, we first define a presheaf model and give the Kripke semantics, and then we show how to interpret the additional type and term formers of guarded recursive type theory.
%%Finally, we conclude that the syntax is consistent in \Cref{sec:consistent}.

%\NV{WIP from this point}

%% To compare the strength of these two approaches, we develop denotational semantics of guarded recursive type theory in a type theory with sized types.
%% This shows that guarded recursive type theory can be done via sized types.
%% More specifically, we define a presheaf model of a simple type theory with guarded recursion \cite{BMSS-synthetic}.
%% Simple types are interpreted via the usual presheaf semantics.

%% In contrast, interpreting guarded recursion and guarded recursive types is more challenging.
%% The most common model in the literature is the topos of trees \cite{BMSS-synthetic}.
%% This is a presheaf topos, whose objects are sets indexed by a natural number.
%% We use a slightly different approach: instead of natural numbers, we use sizes.
%% This means that the required interpretations need to be modified.

%% Our contributions are as follows

%% In Section 2, we describe the meta-theory in which we work.
%% In Section 3, we describe the syntax of the object theory, which is basically a variant of guarded recursion as done by Atkey and McBride \cite{atkey2013productive}.
%% We do this in the form of a deep embedding.
%% More specifically, we define data types representing the contexts, types, terms, and so on, and we give the rules for definitional equality.
%% In Sections 4 and 5, we define the denotational semantics of this syntax.
%% First, we give the interpretation of context, types, and terms in the model, and we show how to interpret simple type formers.
%% Then we show this model supports guarded recursion and guarded recursive types.
%% Finally, we show the syntax is consistent in Section 6.

%% \subsection{Related Work}
%% 
%% Guarded recursion via the later modality was introduced by Nakano \NV{Citation}.

>>>>

% easychair.tex,v 3.5 2017/03/15

\documentclass{easychair}
%\documentclass[EPiC]{easychair}
%\documentclass[EPiCempty]{easychair}
%\documentclass[debug]{easychair}
%\documentclass[verbose]{easychair}
%\documentclass[notimes]{easychair}
%\documentclass[withtimes]{easychair}
%\documentclass[a4paper]{easychair}
%\documentclass[letterpaper]{easychair}

\usepackage{doc}

% use this if you have a long article and want to create an index
% \usepackage{makeidx}

% In order to save space or manage large tables or figures in a
% landcape-like text, you can use the rotating and pdflscape
% packages. Uncomment the desired from the below.
%
% \usepackage{rotating}
% \usepackage{pdflscape}

% Some of our commands for this guide.
%
\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}

\newcommand{\remove}[1]{}

\newcommand{\ie}{\emph{i.e.,} }
\newcommand{\cf}{\emph{c.f.} }
\newcommand{\eg}{\emph{e.g.,} }
\newcommand{\etal}{\emph{et~al.}}
\newcommand{\etc}{\emph{etc.}}
\newcommand{\wrt}{w.r.t.~}
\newcommand{\resp}{resp.~}
\newcommand{\loccit}{\emph{loc. cit.} }

\newcommand{\GTT}{\textbf{GTT}}

%\makeindex

%% Front Matter
%%
% Regular title as in the article class.
%
\title{Guarded Recursion in Agda via Sized Type}

% Authors are joined by \and. Their affiliations are given by \inst, which indexes
% into the list defined using \institute
%
\author{
Niccol\`o Veltri\inst{1}
\and
Niels van der Weide\inst{2}
}

% Institutes for affiliations are also joined by \and,
\institute{
  Department of Computer Science, IT
  University of Copenhagen, Denmark \\
  \email{nive@itu.dk}
\and
   Institute for Computation and Information Sciences, Radboud University, Nijmegen, The Netherlands \\
   \email{nweide@cs.ru.nl}
 }

%  \authorrunning{} has to be set for the shorter version of the authors' names;
% otherwise a warning will be rendered in the running heads. When processed by
% EasyChair, this command is mandatory: a document without \authorrunning
% will be rejected by EasyChair

\authorrunning{Veltri and van der Weide}

% \titlerunning{} has to be set to either the main title or its shorter
% version for the running heads. When processed by
% EasyChair, this command is mandatory: a document without \titlerunning
% will be rejected by EasyChair
\titlerunning{Guarded Recursion in Agda via Sized Types}

\begin{document}

\maketitle

@@

Guarded recursion \cite{Nakano}
is a different approach where the type system is enhanced with a
modality, called ``later'' and written $\triangleright$, encoding time delay in
types. The later modality comes with a general fixpoint combinator for
programming with productive recursive functions and allows the
specification of guarded recursive types.
These types can be used in combination with clock quantification to define  coinductive types \cite{atkey2013productive,bizjak2016guarded,BahrGM17}. Currently there is no
implementation of a calculus for guarded recursion.
\remove{, \eg the type of guarded
streams satisfying the isomorphism $\mathsf{Str} A \simeq A \times
\triangleright (\mathsf{Str} A)$. The tail of a guarded stream is available only
one time step ahead from now, as evidenced by the application of the
later modality to $\mathsf{Str} A$. Coinductive types can be constructed
from guarded recursive types in type theories that additionally
support clock variables and universal quantification over them
\cite{atkey2013productive,BahrGM17}. Currently there is no
implementation of a calculus for guarded recursion.}
%%  nor guarded recursion
%% appears as a selectable feature in a proof assistant.

@@

Utilizing techniques for representing ``type theory in type theory''
\cite{AltenkirchK16,Chapman09}, we present an Agda formalization of
the syntax of a simple type theory for guarded recursion. This object
language, which we call \GTT, is a variant of Atkey and McBride's type
system for productive coprogramming \cite{atkey2013productive} in
which the clock context can contain at most one clock. The types of
\GTT\ include the aforementioned $\triangleright$ modality and a box modality,
a nameless analogue of Atkey and McBride's universal clock
quantification. Clouston \etal \cite{CloustonBGB15} studied a
guarded variant of lambda calculus extended with a box operation,
which they call ``constant''.  Our object calculus differs from
theirs in that our judgments are indexed by a clock context, which can
be empty or containing exactly one clock. 
%In this respect \GTT\ is closer in spirit to Atkey and McBride's
%language.
The design of \GTT\ has the benefit of allowing a more appealing introduction rule for
box than Clouston et al.'s.

@@

This paper only include the essential parts of our Agda
formalization. The full code is available at
\url{https://github.com/niccoloveltri/agda-gtt}. The formalization
uses Agda 2.5.4.2 and Agda standard library 0.16.
\bibliographystyle{plain}
\bibliography{literature}

\end{document}
